---
title: "RNA seq visualizations"
author: "Kesava Asam"
date: "Report Last Run: `r Sys.time()`"
output: 
  pdf_document:
    toc_depth: 2
    number_sections: true
    includes:
      in_header: ~/Documents/Aouizerat_Lab/pdf_req/header.tex
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# copy the logo to the current directory
file.copy("~/Documents/Aouizerat_Lab/pdf_req/alab_logo.png", getwd())
```

# Introduction

RNA seq visualization codes that are used in the Aouizerat lab. The steps are layed out assuming that they are performed on a new project or a new computer. 

# Install packages

```{r}
### CRAN
package_list <- 
c("BiocManager", "tidyverse", "data.table", 
"WriteXLS", "corrplot", "RColorBrewer", "pheatmap", 
"ggrepel", "patchwork", "kableExtra", "ggpubr")

for(package in package_list) {           
  if (!requireNamespace(package, quietly = TRUE))
    install.packages(package)                  
}

### Bioconductor
package_list <- 
c("DESeq2", "apeglm", "EnrichmentBrowser", 
"sva", "RUVSeq", "clusterProfiler", "ReactomePA", 
"org.Hs.eg.db", "OrganismDbi", "Homo.sapiens", "qvalue", 
"airway")

for(package in package_list) {           
  if (!requireNamespace(package, quietly = TRUE))
    BiocManager::install(package)                  
}
```

# Load Packages

```{r results='hide', message=F, warning=F}
library(tidyverse) # Manipulations of datatables
library(airway) # RNA seq data
library(DESeq2) # for DEG generation
library(sva) # for batch corrections
library(RUVSeq) # for RUV batch correction
library(EnrichmentBrowser) # has the new annotation method
library(qvalue) # to determine q or false discovery rate values
library(Homo.sapiens) # for annotating EntrezID to Genesymbols
library(corrplot) # To genearte correlation plots
library(RColorBrewer) # Colors in graphs
library(pheatmap) # for the distance plots
library(ggrepel) # for labeling on graphs without overlap
library(patchwork) # collages of ggplots

library(kableExtra) # for displaying data tables in a pdf
library(ggpubr) # for ggplot collage
```

# Data

```{r}
data("airway")

rna_data <- airway
```

```{r}
ddsSE <- DESeqDataSet(rna_data, design = ~ cell + dex)
ddsSE
```

```{r}
keep <- rowSums(counts(ddsSE)) >= 10
dds <- ddsSE[keep,]
```

```{r}
dds$dex <- relevel(dds$dex, ref = "untrt")
```

```{r}
dds_deseq <- DESeq(dds)
res <- results(dds_deseq)
res
```

```{r}
resultsNames(dds_deseq)
```

```{r}
resLFC <- lfcShrink(dds_deseq, 
                    coef="dex_trt_vs_untrt", 
                    type="apeglm")

resLFC
```

```{r warning=FALSE}
deg_ruv_dat <- as.data.frame(resLFC) %>% 
  rownames_to_column("ENSEMBL")

# use OrganismDb to retrieve gene annotations from homo.sapiens
anno <- AnnotationDbi::select(Homo.sapiens, keytype='ENSEMBL', 
                              keys=deg_ruv_dat$ENSEMBL,
                    columns=c('SYMBOL', 'GENENAME'), multiVals="first")

# Merge the annotation with the results
anno_deg_ruv <- merge(deg_ruv_dat, anno)

### Rearrange the columns for easy viewing
res_deg_ruv <-
anno_deg_ruv %>% 
  arrange(pvalue) %>% 
  dplyr::select(ENSEMBL, SYMBOL, log2FoldChange, pvalue, padj, everything())

res_deg_ruv %>% as_tibble()
```

# PCA plot

Normalize the data using the variance stabilizing method.

```{r}
vsd <- vst(dds, blind=FALSE)
```

A way to visualize sample-to-sample distances is through principal-components analysis (PCA). In this ordination method, the data points (i.e., here, the samples) are projected onto the 2D plane such that they spread out in the two directions which explain most of the differences in the data. The x-axis is the direction (or principal component) which separates the data points the most. The amount of the total variance which is contained in the direction is printed in the axis label.

```{r}
# 1000 genes showing the highest variance are used for PCA
ntop = 1000

# Get the variability of each row
Pvars <- rowVars(assay(vsd))

# Order the Variability and get top genes
select <- order(Pvars, 
                decreasing = TRUE)[
                  seq_len(min(ntop, length(Pvars)))]

# Calculate PCA for the top genes
PCA <- prcomp(t(assay(vsd)[select, ]), scale = F)

# Get the percentage of Variability per PC
percentVar <- round(100*PCA$sdev^2/sum(PCA$sdev^2),1)

# Copy the PCA Data into a Dataframe 
dataGG_nb = data.frame(PC1 = PCA$x[,1], PC2 = PCA$x[,2], 
                    PC3 = PCA$x[,3], PC4 = PCA$x[,4],
                    PC5 = PCA$x[,5], PC6 = PCA$x[,6],
                    sampleNO = colnames(vsd),
                    condition = colData(vsd)$dex)
```

```{r}
# Title of the graph
title <- "Dexamethasone untreated vs treated"

# Caption; data source
caption <- "Data Source: \n Airway Data"

# PCA plot
ggplot(dataGG_nb, aes(x=PC1, y=PC2, 
                     color = condition)) +
  geom_point(size = 2.2, alpha = 0.8) +
  geom_text_repel(data=dataGG_nb, 
             aes(label=sampleNO), size = 3.5) + # Name labels
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  theme_bw() + 
  scale_fill_brewer(type="qual")+
  scale_colour_brewer(type="qual", palette=7) +
  scale_x_continuous(breaks=seq(-70,70,10), limits = c(-25,25)) +
  scale_y_continuous(breaks=seq(-70,70,10), limits = c(-25,25)) +
  labs(title = title, caption = caption, 
       subtitle = "PCs calculated from top 1000 variable genes.")
```

# QQ plot

[QQ plots code](https://slowkow.com/notes/ggplot2-qqplot/)

```{r Func: Q-Q plots}
# Function for creating the qqplots
# Using the Unadjusted P values from results

qq_plot <- function(pvals, ci = 0.95) {
  
  n <- length(pvals)
  df <- data.frame(
  observed = -log10(sort(pvals)),
  expected = -log10(ppoints(n)),
  clower = -log10(qbeta(p = (1 - ci) / 2, shape1 = 1:n, shape2 = n:1)),
  cupper = -log10(qbeta(p = (1 + ci) / 2, shape1 = 1:n, shape2 = n:1))
  )
  log10Pe <- expression(paste("Expected -log"[10], plain(P)))
  log10Po <- expression(paste("Observed -log"[10], plain(P)))
  ggplot(df) +
  geom_ribbon(
  mapping = aes(x = expected, ymin = clower, ymax = cupper),
  alpha = 0.1
  ) +
  geom_point(aes(expected, observed), shape = 1, size = 3) +
  geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
  xlab(log10Pe) +
  ylab(log10Po) +
  theme_bw(base_size = 15) +
  labs(caption = caption, 
       subtitle = "Observed unadjusted P values are plotted against the expected.") +
  annotate(
  geom = "text",
  x = -Inf,
  y = Inf,
  hjust = -0.15,
  vjust = 1 + 0.15 * 3,
  label = sprintf("Inflation Factor = %.2f", inflation(pvals)),
  size = 6.2) +
  theme(
  axis.ticks = element_line(size = 0.5),
  panel.grid = element_blank()
  )
}
```

```{r Func: Calculating lambda}
inflation <- function(pvals) {
chisq <- qchisq(1 - pvals, 1)
lambda <- median(chisq) / qchisq(0.5, 1)
lambda
}
```


```{r}
qq_plot(res_deg_ruv$pvalue)
```


```{r Func: Volcano plot}
#Colour Palette
pal <- c(
  "Down-regulated" = "red",
  "Not significant" = "gray", 
  "Up-regulated" = "blue",
  "~ Down-regulated" = "#E45093",
  "~ Up-regulated" = "#01A5E1"
)

# Function for creating the volcano plots from DEG data frames

vol_plot <- function(df) {
  
  df %>% mutate(key = factor(case_when(padj < 0.05 & log2FoldChange > 1 ~ "Up-regulated", 
                                       padj < 0.05 & log2FoldChange > 0.5 ~ "~ Up-regulated", 
                                       padj < 0.05 & log2FoldChange < -1 ~ "Down-regulated",
                                       padj < 0.05 & log2FoldChange < -0.5 ~ "~ Down-regulated",
                                       TRUE ~ "Not significant"))) %>% 
    ggplot(., aes(x = log2FoldChange, y = -log10(padj), 
                  name = SYMBOL)) +
    geom_point(aes(color = key), size =1.5, alpha = 0.8)+
    scale_color_manual(values = pal)+
    geom_vline(xintercept = 1, color="#b8b8b8", linetype=3)+
    geom_vline(xintercept = -1, color="#b8b8b8", linetype=3)+
    geom_hline(yintercept = -log10(0.1), color="#b8b8b8", linetype=3)+
    scale_x_continuous(breaks=seq(-15,15,1)) +
    scale_y_continuous(breaks=seq(0,200,20)) +
    theme_bw() + 
    labs(subtitle = "-log10 adj P values are represented on the Y axis, and Log 2 Fold change on the X axis.", 
         caption = caption)
  
}
```

```{r}
vol_plot(res_deg_ruv)
```


# Calculate Hidden Batch Effects (SVA)

Calculate Hidden Batch Effects and remove any batches that data may have that we are not aware of

```{r}
dat  <- counts(dds_res, normalized = TRUE)
```

```{r}
idx  <- rowMeans(dat) > 10
dat  <- dat[idx, ]

# Model matrix with all Co-variates and Variable of Interest
mod  <- model.matrix(~dex , data=colData(dds_res))

# Null model matrix with all adjustment variables (Co-variates) 
# Note: This doesnt have the variable of Interest
mod0 <- model.matrix(~1, 
                     data=colData(dds_res))

# Generate the Surrogate Variables
svseq <- svaseq(dat, mod, mod0)
```


### Correlation check

Check to see if the Biological variables and surrogate variables are correlated

Prep Data

```{r}
# Add the surrogate variables as a dataframe
sv_df <- as.data.frame(svseq$sv)

# Rename all the columns
names(sv_df) <- gsub("V", "SV", names(sv_df))

# Get all the Phenotypes of interest
pheno <- 
  colData(dds_res)$dex %>% 
  as.data.frame() %>% setNames("dex") %>% 
  mutate(dex = as.numeric(dex))
  

# Merge the sv_df with he phenotype
pheno_sv <- cbind(pheno, sv_df)
```

Number of significant surrogate variables is:  `r ncol(sv_df)`

Perform correlation to check how the data points are correlated with each other  

```{r}
# Generating the correlation matrix
corr <- cor(pheno_sv)
```

### Visualize

```{r}
col1 <- colorRampPalette(c("#2B226D", "#372C8C", "#4235A9", 
                           "#6259AF", "#B5B1D3",
                           "#D7B49E", "#DC602E", 
                           "#D4592E", "#CC512D", "#BC412B"))

# Generate correlation plot
corrplot(corr, method = "number", 
         type = "full", 
         title = "Correlation between the Phenotypes", mar = c(0, 0, 1, 1),
         col = col1(10),
         number.cex = 0.7,
         number.digits = 2,
         tl.col = "black")
```

